```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Ranking the Harmfulness of Severe Weather Events in the U.S.

### Synopsis

In this analysis, two questions are to be answered for the United States. The 
first question is which types of severe weather events are most harmful with 
respect to public health. The other question is which types of severe weather 
events have the greatest economic consequences. The data to be analysed is from 
the U.S. National Oceanic and Atmospheric Administration (NOAA) database and 
contains the events across the United States from 1950 to 2011. As a result of 
the data analysis, it has been hypothesized that tornado, excessive heat, heat 
and flash flood are the most harmful events to the public health. The other 
hypothesis is that flood, hurricane(typhoon), tornado and storm surge are the most 
harmful evens to the public wealth.

### Data Processing

The analysis data can be obtained from the [storm data][1] in the form of a bzip2
file. The preparation of the data is explained in the [storm data documentation][2].
In addition, there is a document of [frequently asked questions][3] about the 
data.

#### Importing the Raw Data

The data is stored in a comma separated file and the missing values are 
represented by blank fields. The storm data is imported. 

```{r import_data}
if (!file.exists("StormData.csv")){
      the_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
      download.file(the_url, "StormData.csv.bz2")
      the_command <- "bzip2 -d StormData.csv.bz2"
      system(the_command)
}
library(data.table) # load data.table package
library(dplyr) # load dplyr package
# read in data treating "" as NA
data <- fread("StormData.csv", data.table = FALSE, na.strings = "")
# remove duplicate rows in data
data <- distinct(data)
```

#### Processing the Raw Data

After the data is imported, the first few and last few lines of the data are 
displayed.

```{r glance_data}
head(data)
tail(data)
```

A column of the data contains unusually long entries. It is the REMARKS column 
to the left of the REFNUM column. The columns that are of interest are EVTYPE, 
FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. EVTYPE is the 
type of the event. FATALITIES stores the number of the dead people. INJURIES 
gives the number of injured people. PROPDMG is the base part of the property 
damage in terms of dollars. PROPDMGEXP is the exponent part of the property 
damage. CROPDMG and CROPDMGEXP are similar to PROPDMG and PROPDMGEXP with the 
difference that they are for crop damages. The columns that are related with 
the analysis are selected and renamed, the rest are discarded.

```{r narrow_data}
data <- select(data, STATE = STATE__, COUNTY, EVTYPE, FATALITIES, INJURIES, 
               PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

The proportion of NA's is determined.

```{r prop_of_na}
prop_na <- function(x){round(mean(is.na(x)), 2)}
na_table <- sapply(data, prop_na)
print(na_table)
```

The unique values for PROPDMGEXP and CROPDMGEXP are as follows:

```{r dmg_exp}
print(unique(data$PROPDMGEXP))
print(unique(data$CROPDMGEXP))
```

In the storm data preparation manual, it is stated that alphabetical characters 
are used to indicate the magnitude of the damage. They are "K" for thousands, "M" 
for millions and "B" for billions. Hence, the characters in the PROPDMGEXP and 
CROPDMGEXP will be interpreted accordingly. If the exponent character is either 
of "K/k" or "M/m" or "B/b", the damage is multiplied by a thousand or a million or 
a billion, respectively. If the exponent character is a digit, then the magnitude 
is multiplied by 10 to the power being equal to the digit. For all the other 
cases, including NA's, the exponent character has no effect. Hence, the proportions 
of the NA's for PROPDMGEXP and CROPDMGEXP are not important. However, this 
transformation is an approximation to the damage because in the REMARKS column, 
there are some entries giving the amount of damage depending on reports of various 
sources. These damage reports in the REMARKS column are neglected. 

The approximate total damage is calculated and added as a new column to the data.

```{r total_damage}
# function to convert exponent characters to numeric characters
exponent <- function(x){
      if (x %in% c("K", "k")){
            x = "3"
      } else if (x %in% c("M", "m")){
            x = "6"
      } else if (x %in% c("B", "b")){
            x = "9"
      } else if (x %in% c("+", "?", "h", "H", "-")){
            x = "0"
      } else if (is.na(x)){
            x = "0"
      } else {
            x = x
      }
      x
}
# convert property exponent characters to numeric characters
data$PROPDMGEXP <- sapply(data$PROPDMGEXP, exponent)
# convert crop exponent characters to numeric characters
data$CROPDMGEXP <- sapply(data$CROPDMGEXP, exponent)
data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
data <- data %>%
      # compute the property damage and crop damage
      mutate(PROPDMG = PROPDMG * (10 ^ PROPDMGEXP), 
             CROPDMG = CROPDMG * (10 ^ CROPDMGEXP)) %>%
      # discard PROPDMGEXP and CROPDMGEXP columns
      select(STATE, COUNTY, EVTYPE, FATALITIES, INJURIES, 
             PROPDMG, CROPDMG) %>%
      as.data.frame()
# remove extra blanks in the event type names
data$EVTYPE <- trimws(data$EVTYPE)
# convert event type names to lower case
data$EVTYPE <- tolower(data$EVTYPE)
# get the distinct event type names
data_ev_types <- unique(data$EVTYPE)
```

Let the number of unique event types be determined.

```{r unique_events}
num_event_types <- length(data_ev_types)
```

The number of distinct event types is `r num_event_types`. That is a huge number. 
Let the event types given in the storm manual be checked. They are as follows:

```{r def_event_types}
default_event_types <- c("Astronomical Low Tide", "Avalanche", "Blizzard", 
                         "Coastal Flood", "Cold/Wind Chill", "Debris Flow",
                         "Dense Fog", "Dense Smoke", "Drought", "Dust Devil",
                         "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill",
                         "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud",
                         "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow",
                         "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm",
                         "Lake-Effect Snow", "Lakeshore Flood", "Lightning", 
                         "Marine Hail", "Marine High Wind", "Marine Strong Wind",
                         "Marine Thunderstorm Wind", "Rip Current", "Seiche", 
                         "Sleet", "Storm Surge/Tide", "Strong Wind", 
                         "Thunderstorm Wind", "Tornado", "Tropical Depression",
                         "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout",
                         "Wildfire", "Winter Storm", "Winter Weather")
print(default_event_types)
```
There are `r length(default_event_types)` default event types. The number of event
types in the data set is `r num_event_types`. They are very different from each 
other. This difference indicates that there are event type names in the data set 
which are different from the default event type names. Hence, the ranking of the 
event types will be made according to two data sets. First data set is the original 
data set. The second data set will be obtained from the original one by processing 
the event type names in such a way that they are similar to the default event 
type names. Let the default event types and the event types in the data set be 
matched.

```{r process_ev_type}
# default event types are converted to lower case
default_event_types <- tolower(default_event_types)
# "/" character and extra blanks are removed from the default event type names
default_event_types_trimmed <- trimws(gsub("/", " ", default_event_types))
# default event type names are splitted and combined to form the set of default
# type keywords
default_keywords <- unique(unlist(strsplit(default_event_types_trimmed, " ",
                                           fixed = TRUE)))
# "(" character is removed from the default keywords
default_keywords <- gsub("\\(", "", default_keywords)
# ")" character is removed from the default keywords
default_keywords <- gsub("\\)", "", default_keywords)

# clone the original data set
data_modif <- data

# remove the "/" character from the event type names
data_modif$EVTYPE <- gsub("/", " ", data_modif$EVTYPE)
# convert "tstm" to thunderstorm in the event type names
data_modif$EVTYPE <- gsub("tstm", "thunderstorm", data_modif$EVTYPE)
# remove the extra blank spaces in the event type names
data_modif$EVTYPE <- trimws(data_modif$EVTYPE)

# split the event type names into words
splitted_names <- strsplit(data_modif$EVTYPE, " ")

# take the intersection of the splitted names with the default keywords to 
# obtain a name more similar to the default type name
edit_type <- function(x){
      paste(dplyr::intersect(x, default_keywords), collapse = " ")
}

# obtain the names similar to the default event type names
similar_names <- sapply(splitted_names, edit_type)
# update the event type names
data_modif$EVTYPE <- similar_names
```

The processing of the raw data has been completed.

### Results

The processed data is to be analysed to determine which types of severe weather 
events are most harmful with respect to public health and which types of severe 
weather events have the greatest economic consequences. The analyses will be 
performed for the original data set and the data set with the modified event type 
names.

#### Damage to the Public Health

The effect of the events to the public health is to be determined first. The total 
number of deaths and injuries will be found and the events will be ordered with 
respect to death first and injuries second.

```{r public_health_orig_data}
# using the original data set, total fatalities and injuries are computed and 
# data is ordered with respect to decreasing fatalities and injuries
health <- data %>% group_by(EVTYPE) %>% 
      summarise(fatalities = sum(FATALITIES), injuries = sum(INJURIES)) %>%
      arrange(desc(fatalities), desc(injuries)) %>%
      as.data.frame()
head(health, n = 10)
```

The data set with the modified event names is also processed to determine the 
events which are most harmful to the public health.

```{r public_health_modif_data}
# using the modified data set, total fatalities and injuries are computed and 
# data is ordered with respect to decreasing fatalities and injuries
health <- data_modif %>% group_by(EVTYPE) %>% 
      summarise(fatalities = sum(FATALITIES), injuries = sum(INJURIES)) %>%
      arrange(desc(fatalities), desc(injuries)) %>%
      as.data.frame()
head(health, n = 10)
```

From these two results, it is hypothesized that tornado, excessive heat, heat 
and flash flood are the most harmful events to the public health although the 
orders of the heat and flash flood are interchanged in the original data set and 
the modified data set.

The events most harmful to the public health are plotted.

```{r health_harm_plot}
library(ggplot2) # load the package ggplot2
health <- health[1:10, ] # first ten rows are selected
# order the bars with respect to decreasing fatalities 
p <- ggplot(health, aes(x = reorder(EVTYPE, -fatalities), y = fatalities))
p <- p + geom_bar(stat = "identity")
x_label <- "event type" # x axis label
the_caption <- "Fig.1 Events which are most harmful to the public health"
p <- p + labs(x = x_label, caption = the_caption)
# x axis tick labels are made vertical due to space restrictions
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```

#### Damage to the Public Wealth

The damage of the events to the public wealth is to be examined. The total 
expense of the property and the crop damages will be computed and arranged in a 
decresing order.

```{r public_wealth_orig_data}
# using the original data set, total wealth damage is computed and 
# data is ordered with respect to decreasing wealth damage
wealth <- data %>% group_by(EVTYPE) %>% 
      summarise(wealth_damage = sum(PROPDMG) + sum(CROPDMG)) %>%
      arrange(desc(wealth_damage)) %>%
      as.data.frame()
head(wealth, n = 10)
```

The same examination is to be carried out for the data set with the modified 
event type names.

```{r public_wealth_modif_data}
# using the modified data set, total wealth damage is computed and 
# data is ordered with respect to decreasing wealth damage
wealth <- data_modif %>% group_by(EVTYPE) %>% 
      summarise(wealth_damage = sum(PROPDMG) + sum(CROPDMG)) %>%
      arrange(desc(wealth_damage)) %>%
      as.data.frame()
head(wealth, n = 10)
```

From these two results, it is hypothesized that flood, hurricane(typhoon), 
tornado and storm surge are the most harmful evens to the public wealth.

The events most harmful to the public wealth are plotted.

```{r wealth_harm_plot}
wealth <- wealth[1:10, ] # first ten rows are selected
# order the bars with respect to decreasing wealth damage
p <- ggplot(wealth, aes(x = reorder(EVTYPE, -wealth_damage), y = wealth_damage))
p <- p + geom_bar(stat = "identity")
x_label <- "event type" # x axis label
y_label <- "wealth damage" # y axis label
the_caption <- "Fig.2 Events which are most harmful to the public wealth"
p <- p + labs(x = x_label, y = y_label, caption = the_caption)
# x axis tick labels are made vertical due to space restrictions
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "storm data"
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "storm data documentation"
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf "frequently asked questions"
